Exploring Governement government's finances 
========================================================

In order to understand the data i had at hand I started with the first table 
reading the documentaiton.
Table t1 explains "how much money does Uncle Sam have in his checking account on any given day""

I wanted to build a simple cumulative view of the the amount of money in each account at the beginnig fo the fiscal year for the time frame we had at our disposal.

```{r}
library(sqldf)
library(ggplot2)
library(colorspace)
library(scales)
setwd('/Users/carlotorniai/Dropbox/Zipfian/git/government-shutdown')
# Load all of the tables into R
.tables <- sqldf('SELECT tbl_name FROM sqlite_master', dbname = 'treasury_data.db')$tbl_name
for (.table in .tables) {
  if (!(.table %in% ls())) {
    .sql <- paste('SELECT * FROM', .table)
    assign(.table, sqldf(.sql, dbname = 'treasury_data.db'))
  }
}
# Let's try to experiment with some queries..
print('Min date in T1')
print(sqldf('SELECT MIN(date) FROM t1;'))
print('Max date in T1')
print(sqldf('SELECT MAX(date) FROM t1;'))


brewerplot <- function (palette) {
p + scale_fill_brewer(palette = palette) + opts(title=palette)
}
```


Let's see the trend of the amount of money in the Federal Reserve account from 2008
```{r}

# Let's get the data first
sql <- 'select * from t1 where account ="Federal Reserve Account" and date > "2008-01-01"'
federal_account_trend = sqldf(sql)
# Let's transform the dates in date
federal_account_trend$date <- as.Date(federal_account_trend$date)
# Let's plot and ad some axis
ggplot(federal_account_trend, aes(x = as.Date(date), y=open_today, color=account)) + geom_line() + xlab("Year") + ylab("Amount in Milions of US dollars") + labs(title = "Money in the Federal Reserve Account") +  scale_y_continuous(labels = dollar)
```

What I wantd to plot next was the amount of money per type of account at the beginning of each fiscal year from 2005 through 2013.

```{r}
accounts <- sqldf('select distinct account, open_fy, year from t1 where is_total==0 and account <>"Total Opening Balance"')
p1 <- ggplot(accounts, aes(x=year, y=open_fy, group=account, fill=account)) + geom_bar(position="dodge", stat="identity") + xlab("Year") + ylab("Amount in Milions of US dollars") + labs(title = "Opening balance by account type between 2005 and 2013")

p2 <- p1 +  scale_y_continuous(labels = dollar) + scale_x_continuous(breaks=2005:2013)
# Here I want to try to use the scale on Y to display the dollar
# Now I want to add each entry 
p2
```

We can notice during the 2008-2011 the start fo the Supplementing financing program
and the drop in the Tax and Loan Note accounts.

For more info see this:
(Supplementing financing program)[http://www.pyramis.com/fileadmin/templates/pyramis_public/downloads/us/unwinding_the_SFP_WP_Nov_2010.pdf]

I was also interested in seeing more closeley the trend of the Tax and Loan Note Account. This is an account in a "private-sector depository institution, held in the name of the district Federal Reserve Bank as fiscal agent of the United States, that serves as a repository for operating cash available to the U.S. Treasury. Withheld income taxes, employers’ contributions to the Social Security fund, and payments for U.S. government securities routinely go into a tax and loan account.""

("An account at a bank where a Federal Reserve bank deposits taxes that it receives from individual and corporate taxpayers. This increases the liquidity of the banks with the TT&L accounts; this in turn keeps the banking system stable.")[http://www.answers.com/topic/tax-and-loan-account#ixzz2i839QzWQ]


```{r}
sql <- 'select * from t1 where account ="Tax and Loan Note Accounts ( Table V )" and date > "2008-01-01"'
tax_account_trend = sqldf(sql)

# Let's plot and ad some axis
ggplot(tax_account_trend, aes(x = as.Date(date), y=open_today, color=account)) + geom_line() + xlab("Year") + ylab("Amount in Milions of US dollars") + labs(title = "Money in the Tax and Loan Note Accounts") +  scale_y_continuous(labels = dollar)

```

At this point I was curious to see how the money were spent during the financial crisis in particular looking for any significnat change in the expenses before and after 2008.

```{r}
expenses_2005 <- sqldf('select  year, account, item, sum(today) as total from t2 where is_total==0 and year=2005 and transaction_type=="withdrawal" and item not in ("Public Debt Cash Redemp ( Table III B )", "Transfers to Federal Reserve Account Table V", "Transfers to Depositaries") group by account, item order by fytd desc limit 10' )

expenses_2006 <- sqldf('select  year, account, item, sum(today) as total from t2 where is_total==0 and year=2006 and transaction_type=="withdrawal" and item not in ("Public Debt Cash Redemp ( Table III B )", "Transfers to Federal Reserve Account Table V", "Transfers to Depositaries") group by account, item order by fytd desc limit 10' )

expenses_2007 <- sqldf('select  year, account, item, sum(today) as total from t2 where is_total==0 and year=2007 and transaction_type=="withdrawal" and item not in ("Public Debt Cash Redemp ( Table III B )", "Transfers to Federal Reserve Account Table V", "Transfers to Depositaries") group by account, item order by fytd desc limit 10' )

expenses_2008 <- sqldf('select year, account, item, sum(today) as total from t2 where is_total==0 and year=2008 and transaction_type=="withdrawal" and item not in ("Public Debt Cash Redemp ( Table III B )", "Transfers to Federal Reserve Account Table V", "Transfers to Depositaries") group by account, item order by fytd desc limit 10' )

expenses_2009 <- sqldf('select year, account, item, sum(today) as total from t2 where is_total==0 and year=2009 and transaction_type=="withdrawal" and item not in ("Public Debt Cash Redemp ( Table III B )", "Transfers to Federal Reserve Account Table V", "Transfers to Depositaries") group by account, item order by fytd desc limit 10' )

expenses_2010 <- sqldf('select year, account, item, sum(today) as total from t2 where is_total==0 and year=2010 and transaction_type=="withdrawal" and item not in ("Public Debt Cash Redemp ( Table III B )", "Transfers to Federal Reserve Account Table V", "Transfers to Depositaries") group by account, item order by fytd desc limit 10' )

expenses_2011 <- sqldf('select year, account, item, sum(today) as total from t2 where is_total==0 and year=2011 and transaction_type=="withdrawal" and item not in ("Public Debt Cash Redemp ( Table III B )", "Transfers to Federal Reserve Account Table V", "Transfers to Depositaries") group by account, item order by fytd desc limit 10' )

```

```{r}

# Now let's plot this in some meaningful graph let's start with 2008
expenses_2005 <- expenses_2005[complete.cases(expenses_2005),]
#sort_expenses_2008 <- expenses_2008[order(expenses_2008$total, decreasing = TRUE) , ]
expenses_2005 <- transform(expenses_2005, item = reorder(item, total))
exp_2005 <- ggplot(expenses_2005, aes(x=item, y=total, fill=account)) + geom_bar(stat="identity") + xlab("Expense Type") + ylab("Amount in Milions of US dollars") + labs(title = "Top Category Expenses in 2005") + theme(axis.text.x = element_text(angle = 90, hjust = 1, size=12)) + theme(axis.text.y = element_text(hjust = 1, size=12)) + scale_y_continuous(labels = dollar) +  coord_flip()
exp_2005

# Now let's plot this in some meaningful graph let's start with 2008
expenses_2006 <- expenses_2006[complete.cases(expenses_2006),]
#sort_expenses_2008 <- expenses_2008[order(expenses_2008$total, decreasing = TRUE) , ]
expenses_2006 <- transform(expenses_2006, item = reorder(item, total))
exp_2006 <- ggplot(expenses_2006, aes(x=item, y=total, fill=account)) + geom_bar(stat="identity") + xlab("Expense Type") + ylab("Amount in Milions of US dollars") + labs(title = "Top Category Expenses in 2006") + theme(axis.text.x = element_text(angle = 90, hjust = 1, size=12)) + theme(axis.text.y = element_text(hjust = 1, size=12)) + scale_y_continuous(labels = dollar) +  coord_flip()
exp_2006


# Now let's plot this in some meaningful graph let's start with 2008
expenses_2007 <- expenses_2007[complete.cases(expenses_2007),]
#sort_expenses_2008 <- expenses_2008[order(expenses_2008$total, decreasing = TRUE) , ]
expenses_2007 <- transform(expenses_2007, item = reorder(item, total))
exp_2007 <- ggplot(expenses_2007, aes(x=item, y=total, fill=account)) + geom_bar(stat="identity") + xlab("Expense Type") + ylab("Amount in Milions of US dollars") + labs(title = "Top Category Expenses in 2007") + theme(axis.text.x = element_text(angle = 90, hjust = 1, size=12)) + theme(axis.text.y = element_text(hjust = 1, size=12)) + scale_y_continuous(labels = dollar) +  coord_flip()
exp_2007

# Now let's plot this in some meaningful graph let's start with 2008
expenses_2008 <- expenses_2008[complete.cases(expenses_2008),]
#sort_expenses_2008 <- expenses_2008[order(expenses_2008$total, decreasing = TRUE) , ]
expenses_2008 <- transform(expenses_2008, item = reorder(item, total))
exp_2008 <- ggplot(expenses_2008, aes(x=item, y=total, fill=account)) + geom_bar(stat="identity") + xlab("Expense Type") + ylab("Amount in Milions of US dollars") + labs(title = "Top Category Expenses in 2008") + theme(axis.text.x = element_text(angle = 90, hjust = 1, size=12)) + theme(axis.text.y = element_text(hjust = 1, size=12)) + scale_y_continuous(labels = dollar) +  coord_flip()
exp_2008

# Now I just ant to have the legend be nicer but for now let's just don't worry about

# Now let's plot this in some meaningful graph let's start with 2009
expenses_2009 <- expenses_2009[complete.cases(expenses_2009),]

expenses_2009 <- transform(expenses_2009, item = reorder(item, total))
exp_2009 <- ggplot(expenses_2009, aes(x=item, y=total, fill=account)) + geom_bar(stat="identity") + xlab("Expense Type") + ylab("Amount in Milions of US dollars") + labs(title = "Top category Expenses in 2009") + theme(axis.text.x = element_text(angle = 90, hjust = 1, size=12)) + theme(axis.text.y = element_text(hjust = 1, size=12)) + scale_y_continuous(labels = dollar) +  coord_flip()
exp_2009


expenses_2010 <- expenses_2010[complete.cases(expenses_2010),]

expenses_2010 <- transform(expenses_2010, item = reorder(item, total))
exp_2010 <- ggplot(expenses_2010, aes(x=item, y=total, fill=account)) + geom_bar(stat="identity") + xlab("Expense Type") + ylab("Amount in Milions of US dollars") + labs(title = "Top Category Expenses in 2010") + theme(axis.text.x = element_text(angle = 90, hjust = 1, size=12)) + theme(axis.text.y = element_text(hjust = 1, size=12)) + scale_y_continuous(labels = dollar) +  coord_flip()
exp_2010

expenses_2011 <- expenses_2011[complete.cases(expenses_2011),]

expenses_2011 <- transform(expenses_2011, item = reorder(item, total))
exp_2011 <- ggplot(expenses_2011, aes(x=item, y=total, fill=account)) + geom_bar(stat="identity") + xlab("Expense Type") + ylab("Amount in Milions of US dollars") + labs(title = "Expenses  in 2011") + theme(axis.text.x = element_text(angle = 90, hjust = 1, size=12)) + theme(axis.text.y = element_text(hjust = 1, size=12)) + scale_y_continuous(labels = dollar) +  coord_flip()
exp_2011

# Now I want to compare 2005 - 2008
expenses_05_06 <- rbind(expenses_2005, expenses_2006) 
expenses_05_06_07 <- rbind(expenses_05_06, expenses_2007)
expenses_05_08 <- rbind(expenses_05_06_07, expenses_2008)
comulative_top_expenses_05_08 <-ggplot(expenses_05_08, aes(x=item, y = total,fill=item)) + geom_bar(stat="identity") +facet_wrap(~ year)  + scale_y_continuous(labels = dollar)  + coord_flip() + xlab("Expense Type") + ylab("Amount in Milions of US dollars") + labs(title = "Top Category Expenses (2005-2008)")
comulative_top_expenses_05_08
# Let's try with the merging still it doesnt' work.
# Let's try to figure this out later 
# Now I want to show all this in the same order 
expenses_08_09 <- rbind(expenses_2008, expenses_2009) 
expenses_08_09_10 <- rbind(expenses_08_09, expenses_2010)
expenses_08_11 <- rbind(expenses_08_09_10, expenses_2011)
comulative_top_expenses <-ggplot(expenses_08_11, aes(x=item, y = total,fill=item)) + geom_bar(stat="identity") +facet_wrap(~ year)  + scale_y_continuous(labels = dollar) + coord_flip()  + xlab("Expense Type") + ylab("Amount in Milions of US dollars") + labs(title = "Top Category Expenses (2008-2011)")
comulative_top_expenses

# I may want to try later on to visualize one on top of the other...

```

SO and now we came to the govrnement shotdown>

Then I want to ask: who didn't get paid with government shoutdown in the first two weeks of October 2013?

Let's try to figure out who didn't withdraw money in the first two weeks.

```{r}
# October 2012
expenses_october_2012 <- sqldf('select  date, account, item, today  from t2 where is_total==0 and year=2012 and transaction_type=="withdrawal" and item not in ("Public Debt Cash Redemp ( Table III B )", "Transfers to Federal Reserve Account Table V", "Transfers to Depositaries") and month=10 and day<16' )
#Let's remove the Nas
expenses_october_2012 <- expenses_october_2012[complete.cases(expenses_october_2012),]
# Also I want to get rid of every value that is 0 or below ) 
expenses_october_2012_pos <- subset(expenses_october_2012, today > 0)
exp_oct_2012 <- ggplot(expenses_october_2012_pos, aes(x=as.Date(date), y=today)) + geom_point(aes(colour=item))  + scale_y_continuous(labels = dollar)  + xlab("Date") + ylab("Amount in Milions of US dollars") + labs(title = "Item paid by the governement Oct 1-15 2012")
exp_oct_2012


# October 2013
expenses_october_2013 <- sqldf('select  date, account, item, today  from t2 where is_total==0 and year=2013 and transaction_type=="withdrawal" and item not in ("Public Debt Cash Redemp ( Table III B )", "Transfers to Federal Reserve Account Table V", "Transfers to Depositaries") and month=10 and day<16' )
#Let's remove the Nas
expenses_october_2013 <- expenses_october_2013[complete.cases(expenses_october_2013),]
# Also I want to get rid of every value that is 0 or below ) 
expenses_october_2013_pos <- subset(expenses_october_2013, today > 0)
exp_oct_2013 <- ggplot(expenses_october_2013_pos, aes(x=as.Date(date), y=today)) + geom_point(aes(colour=item))  + scale_y_continuous(labels = dollar)  + xlab("Date") + ylab("Amount in Milions of US dollars") + labs(title = "Item paid by the governement Oct 1-15 2013")
exp_oct_2013


# September 2013
expenses_sept_2013 <- sqldf('select  date, account, item, today  from t2 where is_total==0 and year=2013 and transaction_type=="withdrawal" and item not in ("Public Debt Cash Redemp ( Table III B )", "Transfers to Federal Reserve Account Table V", "Transfers to Depositaries") and month=9 and day<16' )

#Let's remove the Nas
expenses_sept_2013 <- expenses_sept_2013[complete.cases(expenses_october_2013),]
# Also I want to get rid of every value that is 0 or below ) 
expenses_sept_2013_pos <- subset(expenses_sept_2013, today > 0)
exp_sept_2013 <- ggplot(expenses_sept_2013_pos, aes(x=as.Date(date), y=today)) + geom_point(aes(colour=item)) 
exp_sept_2013


```

It's hard to understand. Let's try to group it.


```{r}
# Now let's group the total for the Month of OCtober 2013
oct_2013_exp_sum <- sqldf('select  date, account, item, sum(today) as total  from t2 where is_total==0 and year=2013 and transaction_type=="withdrawal" and item not in ("Public Debt Cash Redemp ( Table III B )", "Transfers to Federal Reserve Account Table V", "Transfers to Depositaries") and month=10 and day<16 group by item' )

# Now let's group the total for the month of October 2012
oct_2012_exp_sum <- sqldf('select  date, account, item, sum(today) as total  from t2 where is_total==0 and year=2012 and transaction_type=="withdrawal" and item not in ("Public Debt Cash Redemp ( Table III B )", "Transfers to Federal Reserve Account Table V", "Transfers to Depositaries") and month=10 and day < 16 group by item' )

pos_oct_2013 <- subset(oct_2013_exp_sum, total > 0)
pos_oct_2012 <- subset(oct_2012_exp_sum, total > 0)

g_oct_2013 <- ggplot(pos_oct_2013, aes(x=item, y=total)) + geom_bar(stat="identity", aes(fill=item)) + theme(axis.text.x = element_text(angle = 90, hjust = 1, size=12)) + theme(axis.text.y = element_text(hjust = 1, size=12))  + xlab("Date") + ylab("Amount in Milions of US dollars") + labs(title = "Item paid by the governement Oct 1-15 2013")+ coord_flip() 
g_oct_2013


g_oct_2012 <- ggplot(pos_oct_2012, aes(x=item, y=total)) + geom_bar(stat="identity", aes(fill=item)) + theme(axis.text.x = element_text(angle = 90, hjust = 1, size=12)) + theme(axis.text.y = element_text(hjust = 1, size=12)) + xlab("Date") + ylab("Amount in Milions of US dollars") + labs(title = "Item paid by the governement Oct 1-15 2013")+ coord_flip() 
g_oct_2012
```

Well besides the value one thing is clear:
There are "less" items that were paid on the first half of OCtober 2013 wrt the same time frame of the past year.

Let's see which items were left out:
```{r}
exp_2012 <- data.frame(expenses_october_2012) 
items_2012 <- exp_2012$item

exp_2013 <- data.frame(expenses_october_2013) 
items_2013 <- exp_2013$item
lost_items_oct_12_13 <- setdiff(items_2012, items_2013)
lost_items_oct_12_13

```
These are the different items:
 [1] "Emergency Prep Response ( DHS )"    "Federal Crop Ins Corp"             
 [3] "Federal Financing Bank"             "International Monetary Fund"       
 [5] "Treasury Department programs"       "Foreign Military Sales Program"    
 [7] "Commerce Programs"                  "Federal Transit Admin"             
 [9] "Treasury Dept: Claims Payments"     "Agriculture Loan Payments ( misc )"
[11] "Deposit Insurance Fund"             "State Department"  
Well can this just be a coincidence?
Let's compare the first half of the month of semptember and see if we get 
the same significant difference.

```{r}

expenses_september_2012 <- sqldf('select  date, account, item, today  from t2 where is_total==0 and year=2012 and transaction_type=="withdrawal" and item not in ("Public Debt Cash Redemp ( Table III B )", "Transfers to Federal Reserve Account Table V", "Transfers to Depositaries") and month=9 and day<16' )
#Let's remove the Nas
expenses_september_2012 <- expenses_september_2012[complete.cases(expenses_september_2012),]
# Also I want to get rid of every value that is 0 or below ) 
expenses_september_2012_pos <- subset(expenses_september_2012, today > 0)

# Let's check Septmber 2013
expenses_september_2013 <- sqldf('select  date, account, item, today  from t2 where is_total==0 and year=2013 and transaction_type=="withdrawal" and item not in ("Public Debt Cash Redemp ( Table III B )", "Transfers to Federal Reserve Account Table V", "Transfers to Depositaries") and month=9 and day<16' )
#Let's remove the Nas
expenses_september_2013 <- expenses_september_2013[complete.cases(expenses_september_2013),]
# Also I want to get rid of every value that is 0 or below ) 
expenses_september_2013_pos <- subset(expenses_september_2013, today > 0)

# Compare now the first half of September 2012 with September2013
unique(expenses_sept_2013_pos$item)
unique(expenses_september_2012_pos$item)
```
Well we have 45 and 50. A difference of 5.
Let's see what are the 
```{r}
# See diff between sept 2012 and sept 2013
sept12_13_lost_items <- setdiff(expenses_sept_2012_pos$item, expenses_september_2013_pos$item)
sept12_13_lost_items
```
[1] "Net Change in Operating Cash Balance" "Federal Aviation Administration"     
[3] "Federal Crop Ins Corp"                "International Monetary Fund"         
[5] "Small Business Administration"        "Treasury Dept: Claims Payments"      
[7] "Agriculture Loan Payments ( misc )"   "Export Import Bank" 

Let's now compare these values with our previous difference
```{r}
setdiff(lost_items_oct_12_13, sept12_13_lost_items)

```

Here we have the actual number fo the 7 voices that weren't paid during the shutdwon
[1] "Emergency Prep Response ( DHS )" "Federal Financing Bank"         
[3] "Treasury Department programs"    "Foreign Military Sales Program" 
[5] "Commerce Programs"               "Federal Transit Admin"          
[7] "Deposit Insurance Fund"          "State Department"   

Let's now plot the daily plot for this 7 in the currrent year
(Then I can also compare wiht the previous year looking at what was going on on OCtober 2013)
```{r}
shutdn_with <- sqldf('select date, today, item from t2 where is_total==0 and year=2013 and transaction_type=="withdrawal" and month in (9,10) and item in ("Emergency Prep Response ( DHS )", "Federal Financing Bank", "Treasury Department programs", "Foreign Military Sales Program", "Commerce Programs", "Federal Transit Admin", "Deposit Insurance Fund","State Department")')

# Let's plot this now.
plot_drop <- ggplot(shutdn_with) + aes(x=as.Date(date), y=today, colour=item) +geom_point() + geom_line() + xlab("Date") + ylab("Amount in Milions of US dollars") + labs(title = "Amount paid for the categories affected by the shutdown in 2013")
plot_drop
```
Now let's see instead what was the trend in the timeframe of last year.
```{r}
shutdn_compare <- sqldf('select date, today, item from t2 where is_total==0 and year=2012 and transaction_type=="withdrawal"and month in (9,10) and item in ("Emergency Prep Response ( DHS )", "Federal Financing Bank", "Treasury Department programs", "Foreign Military Sales Program", "Commerce Programs", "Federal Transit Admin", "Deposit Insurance Fund","State Department")')

# Let's plot this now.
plot_drop_2012 <- ggplot(shutdn_compare) + aes(x=as.Date(date), y=today, colour=item) +geom_line()+ geom_line() + xlab("Date") + ylab("Amount in Milions of US dollars") + labs(title = "Amount paid for the categories affected by the shutdown in 2012")
plot_drop_2012
```

In order to better understamd the trend let's remove the State department.
```{r}
shutdn_compare_no_state <- sqldf('select date, today, item from t2 where is_total==0 and year=2012 and transaction_type=="withdrawal"and month in (9,10) and item in ("Emergency Prep Response ( DHS )", "Federal Financing Bank", "Treasury Department programs", "Foreign Military Sales Program", "Commerce Programs", "Federal Transit Admin", "Deposit Insurance Fund")')

# Let's plot this now.
plot_drop_no_state <- ggplot(shutdn_compare_no_state) + aes(x=as.Date(date), y=today, colour=item) +geom_line() + geom_line() + xlab("Date") + ylab("Amount in Milions of US dollars") + labs(title = "Amount paid for the categories affected by the shutdown in 2012")
plot_drop_no_state
```


